\documentclass{article}

\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{mathpartir}
\usepackage[capitalise, noabbrev]{cleveref}
\usepackage{stmaryrd} % Where \llbracket and \rrbracket are defined

\newcommand\graphalg{\textsf{GraphAlg}}
\DeclareMathOperator{\pickAny}{\mathsf{pickAny}}
\DeclareMathOperator{\cast}{\mathsf{cast}}
\newcommand{\ring}{\texttt{ring}}
\newcommand{\tenv}{\mathcal{T}}
\newcommand{\Ring}{\texttt{Ring}}
\newcommand{\ttypet}{\texttt{type}_{\tenv}}

% Definitions from MATLANG paper
\newcommand{\ML}{\mathsf{MATLANG}}
\newcommand{\transp}[1]{#1^T}
\newcommand{\SyntaxStyle}{\mathsf}
\newcommand{\LetIn}[3]{\SyntaxStyle{let}\ #1=#2\ \SyntaxStyle{in}\ #3}
\newcommand{\IfThenElse}[3]{\SyntaxStyle{if}\ #1\ \SyntaxStyle{then}\ #2\ \SyntaxStyle{else}\ #3}
\newcommand{\rowdom}{\mathbf{1}}
\DeclareMathOperator{\diag}{\mathsf{diag}}
\DeclareMathOperator{\Apply}{\mathsf{apply}}
\DeclareMathOperator{\Inv}{\inv}
\newcommand{\one}{\rowdom}
\newcommand{\dotmin}{\mathbin{\dot-}}
\newcommand{\M}{M}
\newcommand{\reals}{\mathbf{R}}
\newcommand{\cmplx}{\mathbf{C}}
\newcommand{\C}{\mathbf{C}}
\newcommand{\nat}{\mathbf{N}}
\newcommand{\pnat}{\nat_{>0}}
\DeclareMathOperator{\var}{var}
\newcommand{\scm}{\mathcal{S}}
\newcommand{\sizevar}{\mathsf{SizeVars}}
\newcommand{\mname}{\mathsf{MatNames}}
\newcommand{\allinst}[1]{\mathcal{I}_{#1}}
\newcommand{\cmats}{\mathcal{M}}
\newcommand{\false}{\bot}
\newcommand{\true}{\top}
\newcommand{\boolinterp}{\nu}
\DeclareMathOperator{\TC}{TC}

% Definitions from for-MATLANG paper
\newcommand{\ffor}[3]{\texttt{for}\, #1,#2 \texttt{.}\, #3}
\newcommand{\Mvar}{\mathcal{V}}
\newcommand{\I}{\mathcal{I}}
\newcommand{\Sch}{\mathcal{S}}
\newcommand{\ttype}{\texttt{type}}
\newcommand{\ttypes}{\texttt{type}_{\Sch}}
\newcommand{\dom}{\mathcal{D}}
\newcommand{\sem}[2]{\llbracket #1 \rrbracket(#2)}
\newcommand{\Mnam}{\mathcal{M}}
\newcommand{\DD}{\texttt{Symb}}
\newcommand{\size}{\texttt{size}}
\newcommand{\mtr}[1]{\textsf{Mat}[#1]}
\newcommand{\RR}{\mathbb{R}}
\newcommand{\NN}{\mathbb{N}}
\newcommand{\conc}{\texttt{mat}}
\newcommand{\bigstep}[3]{#2(#1)=#3}
\newcommand{\typed}[3]{#1\vdash#2:#3}

\title{GraphAlg Language Specification}
\author{Daan de Graaf}
\date{\today}

\begin{document}
\maketitle

\section{Syntax}
\textbf{TODO: Syntax of the full language}

\section{Typing Rules}
\textbf{TODO: Typing rules of the full language}

\section{Core Syntax}
The \graphalg{} Core language is defined by the grammar:

\begin{align*}
    E ::= & \ M                                    & \text{(matrix variable)}           \\
    |     & \ E^{T}                                & \text{(transpose)}                 \\
    |     & \ \one(E)                              & \text{(one-vector)}                \\
    |     & \ \diag(E)                             & \text{(diagonalizate a vector)}    \\
    |     & \ E_1 \cdot E_2                        & \text{(matrix multiplication)}     \\
    |     & \ \Apply[f](E_1,\ \ldots,E_n)          & \text{(pointwise application)}     \\
    |     & \ \text{for}[b](E_d, E_1,\ \ldots,E_n) & \text{(loop over dimension)}       \\
    %|     & \ \text{for}[b](l, E_1,\ \ldots,E_n)   & \text{(loop over range)}           \\
    |     & \ \pickAny(E)                          & \text{(keep first nonzero in row)} \\
    b ::= & \{ M_1 = E_1,\ \ldots, M_n = E_n \}    & \text{(loop body)}                 \\
    f ::= & (C_1,\ \ldots,C_n) \ e                 & \text{(pointwise function)}        \\
    e ::= & \ C                                    & \text{(scalar variable)}           \\
    % CastDimOp (covered by literal case)
    |     & \ r(l)                                 & \text{(scalar literal)}            \\
    |     & \ e_1 \ \{+,\cdot,-,/,=\} \ e_2        & \text{(scalar arithmetic)}         \\
    |     & \ \cast(r, e)                          & \text{(scalar cast)}               \\
\end{align*}

We use the following notation:
\begin{itemize}
    \item $E$ represents an expression that returns a matrix value, while $e$ represents a scalar value expression
    \item $M$ names a matrix variable (defined in a loop body).
          The scalar equivalent is $C$ (defined in a pointwise function).
    \item $l$ is a literal value such as $42$, $3.14$ or `true'.
    \item $s$ is a semiring identifier such as \texttt{bool}, \texttt{int} or \texttt{real}.
\end{itemize}

\section{Core Typing}
It is possible to write a \graphalg{} Core expression that is valid according to the grammar but is not semantically valid.
For example, it is unclear what the semantics of $A \cdot B$ are if $A$ and $B$ have incompatible dimensions, or if there is no matrix value associated with variable $B$.
We capture such requirements using a \emph{type environment} $\tenv$.
A type environment is a function of the form:

\[
    \tenv(M) = T
\]

which assigns the type $T$ to a matrix variable $M$.
The type of a matrix consists of (1) the number of rows, (2) the number of columns and (3) a semiring.
This yields a function type $\tenv : \Mnam \to \DD \times \DD \times \Ring$, where
$\Mnam$ is a finite set of matrix variables, $\DD$ is a finite set of \emph{size symbols}, and $\Ring$ is the set of supported semirings.

The use of abstract size symbols allows type checking expressions without knowing concrete matrix dimensions.
We denote size symbols by Greek letters $\alpha,\beta,\gamma$.
Unit size $1$ is assumed to be in $\DD$.

\textbf{TODO: More on semirings here}

For constructing modified type environments, we use $\tenv[M := T]$ to denote an instance equivalent to $\tenv$, except that matrix variable $M$ maps to type $T$.

The type of an expression $e$ with respect to an environment $\tenv$ is denoted $\ttype_{\tenv}(e)$.
It is defined according to the type rules below.
We call an expression \emph{well-typed} according to environment $\tenv$ if $\ttype_{\tenv}(e)$ is defined.

\begin{itemize}
    \item $\ttypet(M):= \tenv(M)$, for a matrix variable $M\in\Mnam$;
    \item $\ttypet(\transp{E}):= (\beta,\alpha,R)$ if $\ttypet(E)=(\alpha,\beta,R)$;
    \item $\ttypet(\rowdom(E)):= (\alpha,1)$ if $\ttypet(E)=(\alpha,\beta)$;
    \item $\ttypet(\diag(E)):= (\alpha,\alpha,R)$, if $\ttypet(E)=(\alpha,1,R)$;
    \item $\ttypet(e_1 \cdot e_2):= (\alpha,\gamma,R)$ if  $\ttypet(e_1)=(\alpha,\beta,R)$, and $\ttypet(e_2)=(\beta,\gamma,R)$;
    \item $\ttypet(\Apply[f](E_1,\ldots,E_n)) = (\alpha,\beta,R_f)$ if $R_f$ is the output semiring of function $f$ (see below);
    \item $\text{for}[b](E_d, E_1,\ \ldots,E_n) := (\alpha, \beta, R_1)$ if $\ttypet(E_d) = (\gamma, 1, R_d)$, \\$\ttypet(E_1) = (\alpha, \beta, R_1)$ and $b$ is well-typed (see below).
    \item $\ttypet(\pickAny(E)) := T$ if $\ttypet(E) = T$
\end{itemize}

\subsubsection*{Pointwise Function Application}
For type checking pointwise function application $\Apply[f](E_1,\ldots,E_n)$, assume $f = (C_1,\ldots,C_n)\ e$.
then the type environment for the pointwise function is defined as $\tenv_f = \{ C_1 := (1, 1, R_1),\ldots, C_n := (1, 1, R_n) \}$ where $\ttypet(E_i) = (\alpha, \beta, R_i)$ for $1 \leq i \leq n$.
Based on this we can define $\ttype_{\tenv_f}(f) = \ttype_{\tenv_f}(e) = (1, 1, R_f)$.
Typing rules for scalar expressions are given below.
Because scalar expressions always produce a matrix with dimensions $(1, 1)$, we write the type as $R$ instead of $(1, 1, R)$.

\begin{itemize}
    \item $\ttypet(C):= \tenv(C)$, for a matrix variable $C\in\Mnam$;
    \item $\ttypet(r(l)):= r$ if $r \in \Ring$;
    \item $\ttypet(e_1 + e_2):= R$ if $\ttypet(e_1) = \ttypet(e_2) = R$;
    \item $\ttypet(e_1 \cdot e_2):= R$ if $\ttypet(e_1) = \ttypet(e_2) = R$;
    \item $\ttypet(e_1 - e_2):= R$ if $\ttypet(e_1) = \ttypet(e_2) = R$ and $R \in \{ \texttt{int}, \texttt{real} \}$;
    \item $\ttypet(e_1 / e_2):= R$ if $\ttypet(e_1) = \ttypet(e_2) = R$ and $R = \texttt{real}$;
    \item $\ttypet(e_1 = e_2):= \texttt{bool}$ if $\ttypet(e_1) = \ttypet(e_2) = R$;
    \item $\ttypet(\cast(r,e)):= r$ if $r \in \Ring$ and $\ttypet(e) = R$;
\end{itemize}

\subsubsection*{Loop Bodies}
For type checking loop body $b$ in $\text{for}[b](E_d, E_1,\ \ldots,E_n)$ according to environment $\tenv$, assume $b = \{ M_1 := E_1',\ldots,M_n := E_n' \}$.
Let the type environment for the loop body be $\tenv_b = \tenv[M_1 := \ttypet(E_1),\ldots,M_n := \ttypet(E_n)]$.
Then the loop body is well-typed iff $\ttype_{\tenv_b}(E_i') = \ttypet(E_i)$ for all $1 \leq i \leq n$.


% We capture such requirements using a set of type rules and the notion of a \textit{schema}, which for each matrix variable stores 
% (1) the dimensions (how many rows and columns) and (2) the semiring.
% A schema $\Sch$ is a tuple $(\Mnam,\size, \ring)$, where 
% $\Mnam\subset \Mvar$ is a finite set of matrix variables, 
% $\size: \Mnam \to \DD\times \DD$ is a function that maps each matrix variable in $\Mnam$ to a pair of \textit{size symbols}
% and $\ring: \Mnam \to $.
% We denote size symbols by Greek letters $\alpha,\beta,\gamma$. 
% Unit size $1$ is assumed to be in $\DD$.
% The type of an expression $e$ with respect to a schema $\Sch$ is denoted $\ttypes(e)$.
% It is defined according to the following type rules:

% We call an expression \textit{well-typed} according to the schema $\Sch$ if $\ttypes(e)$ is defined.

\section{Core Semantics}
\textbf{TODO: Semantics for operations in the Core Language}

\section{for-$\ML$ with simultaneous induction}
This section defines a variant of for-$\ML$ that matches the expressive power of \graphalg{} Core.

We start by giving the recalling the definitions of $\ML$~\cite{brijder_expressive_2019} and the extension for-$\ML$~\cite{geerts_expressive_2021} that introduces the \emph{canonical for loop} expression.
We then modify for-$\ML$ to include support for multiple loop variables.
We denote the resulting language ga-$\ML$.

Finally, we prove that:
\begin{enumerate}
    \item any \graphalg{} Core expression can be expressed as a ga-$\ML$ expression
    \item any ga-$\ML$ expression can be expressed as a \graphalg{} Core expression
\end{enumerate}

Together, these demonstrate that the languages are equivalent in terms of expressive power.

\subsection{$\ML$}
We start with `vanilla' $\ML$~\cite{brijder_expressive_2019} expressions (without loops), which is defined by the following grammar:
\begin{align*}
    e & ::= \M                                &  & \text{(matrix variable)}             \\
    %&\mid\quad \LetIn{\M}{e_1}{e_2} && \text{(local binding)} \\
      & \mid\quad \transp{e}                  &  & \text{(transpose)}                   \\
      & \mid\quad \rowdom(e)                  &  & \text{(one-vector)}                  \\
      & \mid\quad \diag(e)                    &  & \text{(diagonalization of a vector)} \\
      & \mid\quad e_1 \cdot e_2               &  & \text{(matrix multiplication)}       \\
      & \mid\quad \Apply[f](e_1, \ldots, e_k) &  & \text{(pointwise
        application, $f \in \Omega$)}
\end{align*}

This definition assumes a repertoire $\Omega$ of functions $f : R^k \to R$, representing all possible pointwise functions that can be used with $\Apply$.
The language does not impose any requirements as to which functions should or should not be part of the repertoire.
For our purposes of proving equivalence with \graphalg{} Core we will assume a particular definition of $\Omega$, see \cref{sec:pointwise}.
The definition of $\ML$ in \cite{brijder_expressive_2019} also includes a let binding expression $\LetIn{\M}{e_1}{e_2}$ for convenience.
As noted in \cite{brijder_expressive_2019}, this operation can be omitted without affecting the expressive power of the language, so we omit it from our definition.

\subsubsection*{Typing}
\textbf{TODO: dedupe with typing rules in core}

\textbf{TODO: define $\Mvar$}

It is possible to write a $\ML$ expression that is valid according to the grammar but is not semantically valid.
For example, it is unclear what the semantics of $A \cdot B$ are if $A$ and $B$ have incompatible dimensions, or if there is no matrix value associated with variable $B$.
We capture such requirements using a set of type rules and the notion of a \textit{schema} that associates matrix variables with matrix sizes.
A schema $\Sch$ is a pair $(\Mnam,\size)$, where $\Mnam\subset \Mvar$ is a finite set of matrix variables, and $\size: \Mnam \to \DD\times \DD$ is a function that maps each matrix variable in $\Mnam$ to a pair of \textit{size symbols}.
We denote size symbols by Greek letters $\alpha,\beta,\gamma$.
Unit size $1$ is assumed to be in $\DD$.
The type of an expression $e$ with respect to a schema $\Sch$ is denoted $\ttypes(e)$.
It is defined according to the following type rules:

\begin{itemize}
    \item $\ttypes(M):= \size(M)$, for a matrix variable $M\in\Mnam$;
    \item $\ttypes(\transp{e}):= (\beta,\alpha)$ if $\ttypes(e)=(\alpha,\beta)$;
    \item $\ttypes(\rowdom(e)):= (\alpha,1)$ if $\ttypes(e)=(\alpha,\beta)$;
    \item $\ttypes(\diag(e)):= (\alpha,\alpha)$, if $\ttypes(e)=(\alpha,1)$;
    \item $\ttypes(e_1 \cdot e_2):= (\alpha,\gamma)$ if  $\ttypes(e_1)=(\alpha,\beta)$, and $\ttypes(e_2)=(\beta,\gamma)$;
    \item $\ttypes(\Apply[f](e_1,\ldots ,e_k)):= (\alpha,\beta)$, if $\ttypes(e_i) = (\alpha, \beta)$ for all $i \in [1 \ldots k]$ and $f \in \Omega$.
\end{itemize}

We call an expression \textit{well-typed} according to the schema $\Sch$ if $\ttypes(e)$ is defined.

\subsubsection*{Semantics}
We now proceed by defining the semantics of well-typed expressions.
We use the notation $\mtr{\RR}$ for the set of all real matrices.
For $A\in\mtr{\RR}$, $\dim(A)\in\NN^2$ denotes its dimensions.
The semantics of a $\ML$ expression depends on the concrete sizes and values of the matrix variables.
We define a function $\dom : \DD \to \NN$ which maps size symbols to concrete sizes,
and another function $\conc : \Mnam \mapsto \mtr{\RR}$ to map matrix variables to concrete matrix values.
$\dom$ and $\conc$ are defined such that for a matrix $M$ with $\size(M) = (\alpha, \beta)$, we have $\dim(\conc(M)) = \dom(\alpha) \times \dom(\beta)$.
Together they form an \textit{instance} $I = (\dom, \conc)$ with all the context needed to evaluate an expression that is well-typed according to $\Sch$.
We denote the evaluation of an expression $e$ over instance $I$ as $\sem{e}{\I}$, which we define as follows:

\begin{itemize}
    \item $\sem{V}{\I} := \conc(V)$, for $V\in \Mnam$
    \item $\sem{e^T}{\I} := \sem{e}{\I}^T$, with $A^T$ the transpose of matrix $A$
    \item $\sem{\rowdom(e)}{\I}$ is an $n\times 1$ vector with value $1$ at all positions, where $\dim(\sem{e}{\I})=(n,m)$
    \item $\sem{\diag(e)}{\I}$ is a matrix with the vector $\sem{e}{\I}$ on its main diagonal, and zero in every other position
    \item $\sem{e_1\cdot e_2}{\I} := \sem{e_1}{\I} \cdot \sem{e_2}{\I}$
    \item $\sem{\Apply[f](e_1,\ldots ,e_k)}{\I}$ is a matrix $A$ of the same size as $\sem{e_1}{\I}$, and where $A_{ij}$ has the value $f(\sem{e_1}{\I}_{ij},\ldots ,\sem{e_k}{\I}_{ij})$
\end{itemize}

\subsection{Canonical Loop Definition}
for-$\ML$~\cite{geerts_expressive_2021} adds the \emph{canonical for loop} expression:

\medskip
\begin{tabular}{lc}
    $\ffor{v}{X}{e}$ & (canonical for loop, with $v, X \in \Mvar$).
\end{tabular}
\medskip

\noindent
This loop construct repeatedly updates state variable $X$ by evaluating $e$, where $e$ is allowed to refer to the previous state of the loop as $X$.
Additionally, $e$ may refer to the \emph{canonical vector} $v$.
Assume a set of canonical vectors $b_1^n,\ldots,b_n^n$ of dimension $n$ defined as $b_1^n = \transp{[1\ 0 \cdots 0]},b_2 = [0\ 1\ 0 \cdots 0]^T$.
In the first iteration, $v$ takes on value $b_1^n$, in the second iteration $b_2^n$, etc.

Following~\cite{geerts_expressive_2021}, we introduce the following typing rule for the added loop construct:

\begin{itemize}
    \item $\ttypes(\ffor{v}{X}{e}) := (\alpha,\beta)$, if \\
          $\ttypes(e)=\ttypes(X) =(\alpha,\beta)$ and $\ttypes(v) = (\gamma,1)$.
\end{itemize}

Assuming $n := \dom(\gamma)$, $\sem{\ffor{v}{X}{e}}{\I}$ is defined iteratively as follows:
\begin{itemize}
    \item Let $A_0 := \mathbf{0}$ be the zero matrix of size $\dom(\alpha)\times \dom(\beta)$.
    \item For $i=1,\ldots n$, compute $A_i:= \sem{e}{\I[v := b_i, X:= A_{i-1}]}$.
    \item Finally, set $\sem{\ffor{v}{X}{e}}{\I}:= A_{n}$.
\end{itemize}

The notation $\I[M := A]$ denotes an instance equivalent to $\I$, except that matrix variable $M$ maps to value $A$.

\subsection{Adding Simultaneous Induction}
\textbf{TODO: Motivate why do we need simultaneous induction}

ga-$\ML$ starts from for-$\ML$, but modifies the loop definition to allow multiple loop variables:

\medskip
\begin{tabular}{lc}
    $\texttt{for}\, v,\{ M_1 := e_1', \ldots, M_m := e_m' \}(e_1, \ldots, e_m)$
\end{tabular}
\medskip

The loop body defines a number of loop variables $m \geq 1$ that each consist of a \emph{binding variable} $M_i$ and a \emph{body expression} $e_i'$.
Intuitively, before entering the loop, the binding variables are initialized as $M_i := e_i$.
Evaluation of the loop is similar to for-$\ML$, each time updating the binding variables with the results of the body expressions.
The loop produces a single result matrix, which is the final value of $M_1$.
Expression $e_d$ is used to define the number of loop iterations $n$.

The typing rule is modified accordingly:

\begin{itemize}
    \item $\ttypes(\texttt{for}\, v,\{ M_1 := e_1', \ldots, M_m := e_m' \}(e_d, e_1, \ldots, e_m)) := \tau_1$,\\
          if $\tau_i := \ttypes(e_i) = \ttype_{\Sch'}(e_i')$ for all $i \in [1 \cdots m]$,\\
          where $\Sch' := \Sch[M_1 := \tau_1,\ldots,M_m := \tau_m]$
          and $\ttypes(v) = (\gamma,1)$.
\end{itemize}

The notation $\Sch[M := \tau]$ denotes a schema equivalent to $\Sch$, except that matrix variable $M$ maps to type $\tau$.

Assume that $\ttypes(v) = (\gamma,1)$.
Then we assign $n := \dom(\gamma)$.
The semantics of the loop are defined iteratively, as follows:
\begin{itemize}
    \item Let $\I_0 := \I[v := b^{n}_0, M_1 := \sem{e_1}{\I}, \ldots, M_m := \sem{e_m}{\I}]$
    \item For $i = 1,\ldots n$, compute $\I_i := \I_{i-1}[v := b^{n}_i, M_1 := \sem{e_1'}{\I_{i-1}}, \ldots, M_m := \sem{e_m'}{\I_{i-1}}]$.
    \item Finally, set $\sem{\texttt{for}\, v,\{ M_1 := e_1', \ldots, M_m := e_m' \}(e_1, \ldots, e_m)}{\I}:= \I_n(M_1)$.
\end{itemize}

The ga-$\ML$ loop definition subsumes the for-$\ML$ version. A for-$\ML$ loop can be simulated as follows:

\begin{tabbing}
    L0\=L1\=L2\=L3\=L4 \kill
    $\ffor{v}{X}{e} \Rightarrow$\\
    $\texttt{for}\, v,\{ X := e \}(\Apply[0](X))$\\
\end{tabbing}

We use $\Apply[0](X)$ to create a matrix with the shape of $X$ with value $0$ at all positions.

\subsection{Pointwise Functions}\label{sec:pointwise}
The syntax definition of $\ML$ assumes a repertoire $\Omega$ of functions $f : R^k \to R$, representing all possible pointwise functions that can be used with $\Apply$.
Here $R$ is a semiring, often the complex or real numbers.
The definition of $\ML$ does not dictate a particular set of functions in $\Omega$.
Instead, $\Omega$ is chosen depending on the required scalar functions for particular proofs.

In \graphalg{} the set of allowed scalar operations is made explicit in the syntax.
In the following sections, we assume $\Omega$ to comprise the set of all possible functions that can be constructed by combining the scalar operations in \graphalg{}.

\subsection{On Casting Between Semirings}
It is typically assumed that all expressions in a $\ML$ expression operate over the same semiring.
\graphalg{}, in contrast, introduces a cast operator to switch between different semirings.
This feature is particularly significant because the semiring of a matrix influences the semantics of operations such as matrix multiplication (by changing the addition and multiplication operators) and $\one(e)$ (by changing the identity value $\one$).
While a convenient abstraction, semiring casting does not make \graphalg{} more expressive.
As demonstrated below, all semirings supported in \graphalg{} can be simulated in the semiring $\RR \bigcup \{ \infty \}$, using the regular addition and multiplication operator.

\subsubsection*{Trivial Operations}
A first observation is that the semantics of $\transp{e}$, $\diag(e)$ and loops is independent of the semirings associated with the inner expressions, and hence simulation of such operations is trivial.

\subsubsection*{Pointwise Functions}
The $\Apply$ operation is also largely independent of the semiring used, except that scalar operations within pointwise functions also need to be simulated.

\textbf{TODO: Can we avoid discussing pointwise functions by leaning on our previous statement that `$\Omega$ is assumed to comprise the set of all possible functions that can be constructed by combining scalar operations in \graphalg{}?'}

Simulations for non-trivial scalar operations are given below.
Inner expressions are assumed to have semiring $R$, with addition operator $\oplus$ and multiplication operator $\otimes$.
%Scalar variable references and literals require no changes and are thus trivially simulated.
%Arithmetic operations can be simulated as follows:

\begin{tabbing}
    L0\=L1\=L2\=L3\=L4 \kill
    $e_1 + e_2 \Rightarrow e_1 \oplus e_2$ \\
    $e_1 \cdot e_2 \Rightarrow e_1 \otimes e_2$ \\
    $\cast_R(r,e) \Rightarrow \cast(R, r, e)$ \\
\end{tabbing}

Division is trivial to simulate because it is only defined when $R = \RR$.
Subtraction additionally allows integer inputs, but in this case simulation is also trivial, because integers are included in $\RR$.

\textbf{TODO: Detail how cast op works}

\subsubsection*{Matrix Multiplication}
To rewrite matrix multiplication, we need the following auxiliary function definitions.
Function $\mathsf{emax}(V)$ follows the definition given in \cite{geerts_expressive_2021}, returning the last canonical vector:

\begin{tabbing}
    L0\=L1\=L2\=L3\=L4 \kill
    $\mathsf{emax}(V) = $\\
    $\SyntaxStyle{let}\ v = V \ \SyntaxStyle{in}$ \\
    $\SyntaxStyle{let}\ X = V \ \SyntaxStyle{in}$ \\
    $\texttt{for}\, v,X \texttt{.}\, v$
\end{tabbing}

The converse $\mathsf{emin}$, which returns the first canonical vector, can be defined as $\mathsf{emin}(V) = \transp{\pickAny(\transp{\one(V)})}$.
We also take the definition of ordering matrix $S_{\leq}$ from \cite{geerts_expressive_2021}:

\begin{tabbing}
    L0\=L1\=L2\=L3\=L4 \kill
    $S_{\leq}(V) = \begin{bmatrix}
            1      & 1      & \cdots & 1 \\
            0      & 1      & \cdots & 1 \\
            \vdots & \vdots & \ddots & 1 \\
            0      & 0      & \cdots & 1
        \end{bmatrix} = $\\
    $\SyntaxStyle{let}\ v = V \ \SyntaxStyle{in}$ \\
    $\SyntaxStyle{let}\ X = \diag(V) \ \SyntaxStyle{in}$ \\
    $(\texttt{for}\, v,X \texttt{.}$\\
    \>$X
        + (X \cdot \mathsf{emax}(V) + v) \cdot \transp{v}
        + v \cdot \transp{\mathsf{emax}(V)}
        ) - \one(V) \cdot \transp{\mathsf{emax}(V)}$\\
\end{tabbing}

By subtracting the identity matrix, we obtain $S_{<} = S_{\leq}(V) - \diag(\one(V))$.
Based on $S_{<}$, we can define function $\mathsf{rotate}(V)$:
% NOTE: We could also omit the emax(V) * emin(V).T part to just implement shift.
% Then in sum[..](..) we MUST use (emin(V) * for) rather than emax(V), as we only populate the first position of the vector.

\[
    \mathsf{rotate(V)} = \pickAny(S_{<}(V)) + \mathsf{emax}(V) \cdot \transp{\mathsf{emin}(V)} =
    \begin{bmatrix}
        0      & 1      & 0      & \cdots & 0      \\
        \vdots & \ddots & \ddots & \ddots & \vdots \\
        \vdots &        & \ddots & \ddots & 0      \\
        0      &        &        & \ddots & 1      \\
        1      & 0      & \cdots & \cdots & 0      \\
    \end{bmatrix}
\]

The resulting matrix has the property when it is multiplied with a vector, it rotates all entries by one position.
For example, given $V = \transp{[1\ 2\ 3]}$, we have $\mathsf{rotate}(V) \cdot V = \transp{[2\ 3\ 1]}$.

Function $\mathsf{sum}[\oplus](V)$ takes as input a (column) vector $V$ and returns the sum of all entries in the vector according to addition operator $\oplus$ ($V_0 \oplus V_1 \oplus \ldots$).
It can be defined in for-$\ML$ as:

\begin{tabbing}
    L0\=L1\=L2\=L3\=L4 \kill
    $\mathsf{sum}[\oplus](V) = $\\
    $\SyntaxStyle{let}\ v = V \ \SyntaxStyle{in}$ \\
    $\SyntaxStyle{let}\ X = V \ \SyntaxStyle{in}$ \\
    $\SyntaxStyle{let}\ R = \mathsf{rotate}(V) \ \SyntaxStyle{in}$ \\
    $\transp{\mathsf{emax}(V)} \cdot \texttt{for}\, v,X \texttt{.}\, \Apply[\oplus](V, R \cdot X)$
\end{tabbing}

Finally, matrix multiplication with addition operator $\oplus$ and multiplication operator $\otimes$ can be written as:

\begin{tabbing}
    L0\=L1\=L2\=L3\=L4 \kill
    $A \cdot_{\oplus,\otimes} B \Rightarrow$\\
    $\SyntaxStyle{let}\ v = \one(A) \ \SyntaxStyle{in}$ \\
    $\SyntaxStyle{let}\ w = \one(\transp{B}) \ \SyntaxStyle{in}$ \\
    $\SyntaxStyle{let}\ X = A \cdot B \ \SyntaxStyle{in}$ \\
    $\SyntaxStyle{let}\ Y = A \cdot B \ \SyntaxStyle{in}$ \\
    $\texttt{for}\, v,X \texttt{.}\,$\\
    \>$\SyntaxStyle{let}\ A_r = \transp{v} \cdot A \ \SyntaxStyle{in}$ \\
    \>$X + \texttt{for}\, w,Y \texttt{.}\,$\\
    \>\>$\SyntaxStyle{let}\ B_c = B \cdot w \ \SyntaxStyle{in}$ \\
    \>\>$Y + v \cdot \mathsf{sum}[\oplus](\Apply[\otimes](\transp{A_r}, B_c)) \cdot \transp{w}$
\end{tabbing}

\subsubsection*{One-vector}
Given an expression $e$ with semiring $R$, we can simulate $\one_R(e)$ as $\Apply[1_R](\rowdom(e))$, where $1_R$ is the multiplicative identity of $R$.

\subsubsection*{pickAny}
For $\pickAny$ we only need to account for different definitions of `nonzero'.
Assume input expression $e$ has semiring $R$.
We define the following rewrite rule:

\begin{tabbing}
    L0\=L1\=L2\=L3\=L4 \kill
    $\pickAny_R(e) \Rightarrow \Apply[v,p. v \text{ if } p \text{ else } 0_R]($\\
        \>$e,$\\
        \>$\pickAny(\Apply[\neq 0_R](e)))$
\end{tabbing}

The inner $\Apply$ creates a matrix with value $1$ at positions where $e$ is nonzero according to semiring $R$.
The $\pickAny(..)$ creates a selection mask from this for which values of $e$ to keep.
Finally, that mask is applied by the outer $\Apply$.

\subsection{Simulating ga-$\ML$ in \graphalg{} Core}
We show how ga-$\ML$ expressions can be rewritten into \graphalg{} Core expressions with equivalent semantics.
The two languages share a number of operations. For the following operations, no rewriting is necessary:
\begin{itemize}
    \item matrix variable
    \item transpose
    \item diagonalize a vector
    \item pointwise application.
          Recall our definition of $\Omega$, which means that all pointwise functions are valid in \graphalg{} Core.
    \item matrix multiplication
    \item one-vector
\end{itemize}

This leaves only for loops to be converted, which we can do with the following rewrite rule:

\begin{tabbing}
    L0\=L1\=L2\=L3\=L4 \kill
    $\texttt{for}\, v,\{ M_1 := e_1', \ldots, M_m := e_m' \}(e_1, \ldots, e_m) \Rightarrow$ \\
    $\texttt{for}\, \{$ \\
    \>$v := \pickAny(V - v)$ \\
    \>$V := V - v$ \\
    \>$ M_1 := e_1', \ldots, M_m := e_m' \}($\\
    \>$\pickAny(V),$\\
    \>$\one(v),$\\
    \>$e_1, \ldots, e_m)$ \\
\end{tabbing}

The variable $V$ stores $1$ at all positions for which we have not seen a canonical vector $v$ yet.
At the end of each iteration, one of the positions in $V$ is cleared (by subtracting the current canonical vector $v$).

\subsection{Simulating \graphalg{} Core in ga-$\ML$}
Only two operations from \graphalg{} Core are not present in ga-$\ML$:
\begin{enumerate}
    \item Loops without a canonical vector.
    \item The \texttt{pickAny} operation
\end{enumerate}

We show how to rewrite these expression from \graphalg{} Core into equivalent ga-$\ML$.
For loops, since the semantics of \texttt{for} in \graphalg{} Core are otherwise identical to those of ga-$\ML$, we can trivially rewrite any loop using the rule:
\begin{align*}
    \texttt{for}\ \ \  & \{ M_1 := e_1', \ldots, M_m := e_m' \}(v, e_1, \ldots, e_m)
    \Rightarrow                                                                      \\
    \texttt{for}\, v,  & \{ M_1 := e_1', \ldots, M_m := e_m' \}(e_1, \ldots, e_m)
\end{align*}

$\text{pickAny}(A)$ can be simulated using the for loop with canonical vectors available in for-$\ML$ (and by extension also in ga-$\ML$):

\begin{tabbing}
    L0\=L1\=L2\=L3\=L4 \kill
    $\pickAny(A) \Rightarrow$ \\
    $\SyntaxStyle{let}\ v = \one(A) \ \SyntaxStyle{in}$ \\
    $\SyntaxStyle{let}\ w = \one(\transp{A}) \ \SyntaxStyle{in}$ \\
    $\SyntaxStyle{let}\ X = A \ \SyntaxStyle{in}$ \\
    $\SyntaxStyle{let}\ Y = \transp{w} \ \SyntaxStyle{in}$ \\
    $\SyntaxStyle{let}\ B = w \cdot \transp{w} \ \SyntaxStyle{in}$ \\
    $\texttt{for}\, v,X \texttt{.}\,$\\
    \>$\SyntaxStyle{let}\ R = \transp{v} \cdot A \ \SyntaxStyle{in}$ \\
    \>$X + v \cdot (\texttt{for}\, w,Y \texttt{.}\,$\\
        \>\>$\SyntaxStyle{let}\ D=Y \cdot B\ \SyntaxStyle{in}$ \\
        \>\>$\SyntaxStyle{let}\ P=R \cdot \diag(w)\ \SyntaxStyle{in}$ \\
        \>\>$\Apply[y, d, p.\ \SyntaxStyle{if}\ d = \mathbf{0}\ \SyntaxStyle{then}\ p\ \SyntaxStyle{else}\ y](Y, D, P))$\\
\end{tabbing}

Initial values are assigned to $v$, $w$, $X$, $Y$ (using $\SyntaxStyle{let}\ \ldots \ \SyntaxStyle{in}\ \ldots$) only to define their dimensions, as this is needed for type checking the \texttt{for} loops.
The outer loop $\texttt{for}\, v,X$ iterates over the input rows, while $\texttt{for}\, w,Y$ iterates over the columns.
$Y$ contains the (partial) result for input row $R$.
In the inner loop, we first detect ($D$) if we already have a value for the current row.
This uses the `broadcast' matrix $B$, which broadcasts any nonzero value in $Y$ to all positions in $D$.
Then, we take the proposed ($P$) value at position $(v, w)$ in $A$, setting all other values to zero.
Finally, the inner $\Apply$ either picks the proposed value $p$ if there is no value yet for the current row, or the existing value $y$ otherwise.

\bibliographystyle{plain}
\bibliography{zotero}

\end{document}
